---
title: "Metrics and ML - Data Cleaning and Preprocessing"
author: "Sydney Gu and Isabella Lin"
date: "2025-02-02"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r}
library(dplyr)
library(tidyr)
library(haven)
library(glmnet)
library(knitr)
library(kableExtra)
library(caret)
library(stats)
library(fastDummies)
```

Read in data sets:
```{r}
cex <- read_dta("~/Desktop/metrics_and_ml/ecma31330_final/113270/cexall.dta")
psid <- read_dta("~/Desktop/metrics_and_ml/ecma31330_final/113270/data.dta")
natpr <- read_dta("~/Desktop/metrics_and_ml/ecma31330_final/113270/natpr.dta")
tax9192 <- read_dta("~/Desktop/metrics_and_ml/ecma31330_final/113270/tax9192.dta")
finprc <- read_dta("~/Desktop/metrics_and_ml/ecma31330_final/113270/finprice.dta")
```

Impute missing federal income tax values for observations in 1991 and 1992:
```{r}
psid <- merge(psid, tax9192, by = c("id", "year"), all.x=TRUE)
psid$ftax[psid$year %in% c(1991, 1992)] <- psid$fiitax[psid$year %in% c(1991, 1992)]
psid <- psid[, !(names(psid) %in% c("fiitax"))]
```

Add national price data to PSID data:
```{r}
psid <- merge(psid, natpr, by = "year")
```

Adjust hours worked values in CEX dataframe:
```{r}
cex <- cex %>%
  mutate(hours = weekh * hourh)
cex <- cex %>%
  mutate(hourw = weekw * hourw)
```

Add columns to CEX for federal taxes and assets:
```{r}
cex <- cex %>%
  mutate(
    ftax = income - income_at,  
    asset = pf + pfaw          
  )
```

Convert states in PSID into regions:
```{r}
state_to_region <- function(state_code) {
  if (state_code %in% c(9, 23, 25, 33, 34, 36, 42, 44, 50)) {
    return(1)  # Northeast
  } else if (state_code %in% c(17, 18, 19, 20, 26, 27, 29, 31, 38, 39, 46, 55)) {
    return(2)  # Midwest
  } else if (state_code %in% c(1, 5, 10, 11, 12, 13, 21, 22, 24, 28, 37, 40, 45, 47, 48, 51, 54)) {
    return(3)  # South
  } else if (state_code %in% c(2, 4, 6, 8, 15, 16, 30, 32, 35, 41, 49, 53, 56)) {
    return(4)  # West
  } else {
    return(NA)  # Unknown state code
  }
}

psid <- psid %>%
  mutate(region = sapply(state, state_to_region))
```

Recode PSID race so it matches in CEX:
```{r}
psid <- psid %>%
  mutate(race = case_when(
    race == 1 ~ 1,  # White
    race == 2 ~ 2,  # Black or African American
    race == 3 ~ 3,  # Native American
    race == 4 ~ 4,  # Asian
    race == 5 ~ NA, # Latino origin
    race == 6 ~ NA, # Other color
    race == 7 ~ NA, # Other
    race == 9 ~ NA, # Don't know
    TRUE ~ NA      
  ))
```


Standardize household income and food expenditure in PSID:
```{r}
psid <- psid %>%
  mutate(income = ly + wly + y) %>%
  mutate(food = food + fout)
```

Remove unnecessary columns in CEX:
```{r}
cex <- cex %>% select(-c(complete, cons1, cons2, cpi, educw, family, famtype, income, income_net, pf, pfaw, pfin, wageh, wagew, weekh, weekw, yb, ndurplus))
```


Rename the variables that have the same values across both datasets:
```{r}
cex <- cex %>%
  rename(fsize = ncomp,
         kids = kid,
         educ = educh,
         income = income_at,
         )
```

Convert all columns in CEX to numeric:
```{r}
cex <- cex %>%
  mutate(across(everything(), ~ as.numeric(as.character(.))))
```

Convert husband and wife hours worked columns to 0 in place of NA:
```{r}
cex <- cex %>%
  mutate(across(c(hours, hourw), ~ ifelse(is.na(.), 0, .)))
```

Remove observations with negative income in CEX and PSID:
```{r}
cex <- cex[cex$income >= 0, ]
psid <- psid[psid$income >= 0, ]
```


Create dummy columns for categorical variables in PSID:
```{r}
#psid <- dummy_cols(psid, select_columns = c("race", "region"))
#psid <- psid %>%
 # select(-c(race, region, state))
```

Create dummy columns for categorical variables in CEX:
```{r}
#cex <- dummy_cols(cex, select_columns = c("race", "region"))
#cex <- cex %>%
 # select(-c(race, region))
```

