---
title: "Metrics and ML - Data Cleaning and Preprocessing"
author: "Sydney Gu and Isabella Lin"
date: "2025-02-02"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r}
library(dplyr)
library(tidyr)
library(haven)
library(glmnet)
library(knitr)
library(kableExtra)
library(caret)
library(stats)
```

```{r}
cex <- read_dta("~/Desktop/metrics_and_ml/ecma31330_final/113270/cexall.dta")
psid <- read_dta("~/Desktop/metrics_and_ml/ecma31330_final/113270/data.dta")
```

Create dummy columns for categorical variables in PSID:
```{r}
psid <- dummy_cols(psid, select_columns = c("race", "state", "smsa"))

# Remove columns that represent >2 categorical variables
psid <- psid %>%
  select(-c(vet, disab, outkid, unioni, unionj, race, state, smsa))
```

Create dummy columns for categorical variables in CEX:
```{r}
cex <- dummy_cols(cex, select_columns = c("race", "region", "famtype"))

# Remove columns that represent >2 categorical variables
cex <- cex %>%
  select(-c(race, region, famtype))
head(cex)
```

```{r}
natpr <- read_dta("~/Desktop/metrics_and_ml/ecma31330_final/113270/natpr.dta")
tax9192 <- read_dta("~/Desktop/metrics_and_ml/ecma31330_final/113270/tax9192.dta")
finprc <- read_dta("~/Desktop/metrics_and_ml/ecma31330_final/113270/finprice.dta")
```

Impute missing federal income tax values for observations in 1991 and 1992:
```{r}
psid <- merge(psid, tax9192, by = c("id", "year"))
psid$ftax[psid$year %in% c(1991, 1992)] <- psid$fiitax[psid$year %in% c(1991, 1992)]
psid <- psid[, !(names(psid) %in% c("fiitax"))]
```

Add national price data to PSID data:
```{r}
psid <- merge(psid, natpr, by = "year")
```

Remove unnecessary columns:
```{r}
cex <- cex %>% select(-c(complete, cons1, cons2, family))
```

Adjust hours worked values in CEX dataframe:
```{r}
cex <- cex %>%
  mutate(hours = weekh * hourh)
cex <- cex %>%
  mutate(hourw = weekw * hourw)
```

Standardize household income variables:
```{r}
# for PSID 
psid <- psid %>%
  mutate(total_income = tyhw + tyoth)
```

```{r}
psid <- psid %>%
  mutate(hw_laborinc = ly + wly)
```

```{r}
cex <- cex %>%
  mutate(hw_wages = wageh + wagew)
```

```{r}
psid <- psid %>%
  mutate(hwage_inc = wages * hours)
```

Now, let's rename the variables that have the same values across both datasets:
```{r}
cex <- cex %>%
  rename(fsize = ncomp,
         kids = kid,
         educ = educh,
         total_income = income_at,
         )
```

Convert all columns in CEX to numeric:
```{r}
cex <- cex %>%
  mutate(across(everything(), ~ as.numeric(as.character(.))))
```

Convert husband and wife columns to 0 in place of NA:
```{r}
cex <- cex %>%
  mutate(across(c(hourh, hourw, wageh, wagew, hours, hw_wages), ~ ifelse(is.na(.), 0, .)))
```
